Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-06-08 analysis/paper/paper.qmd
updates

```{r}
# 读取 Excel 文件
library(tidyverse)
library(here)
library(readxl)

data_clean <- data %>% 
  drop_na()

n_of_artefacts <- nrow(data_clean)
```

Based on typological study, we identify `r n_of_artefacts` rounded faceted sub-spheroids/spheroids in QSY site. Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers. \[BM: feels like this belongs in results section\]

~~To avoid preconceptions and self-justification,~~ We ~~try to~~ applied quantitative methods to ~~help confirm the spheroids in site. These morphological measurements designed to~~ indicate ~~whether these identified spheroids are spherical and~~ the degree of difference between the archaeological spheroids and a perfect sphere. This included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy and the centroid offset for each artefact. We ~~will~~ also examined the degree of standardization of the spheroids by comparing coefficients of variation \[citation\] of sphericity \[and what else\].

@@

Where V is the volume and A is the surface area of the object. This dimensionless metric yields a maximum value of 1 for a perfect sphere. Values closer to 1 indicate a higher degree of sphericity, thus serving as a reliable proxy for assessing the morphological regularity and intentional shaping of the spheroids. We used Python to derive surface area and volume measurements from 3D mesh models. The models were were decimated to a consistent face count and subjected to unit normalization, ensuring comparability across specimens. \[BM: check if we need this here\]

@@

Spherical Harmonic (SPHARM) analysis is a method for describing 3D shapes using a series of mathematical functions defined on the surface of a sphere. It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies. Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$. The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes. Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively. Lower-order terms describe the general shape, while higher-order terms capture finer details[@Link2024; @Medyukhina2020; @Harper2022]. This method is widely used in the field of biology, such as ~~for~~ the analysis of skull morphology and cell morphology \[citations\]. However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025]. Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.

We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) \[citation\]. Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme \[citation\]. The resulting coefficients were normalized for scale-invariant comparison across models.

@@

### Spherical Harmonic energy

@@

To further investigate technological strategies, we incorporated orientation statistics as a complementary approach. This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts. We followed the protocol developed by Sam Lin [@Lin2024], extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps are finished in Geomagic Wrap, Rhino8 and R.

@@

Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies. We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes. This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution. A total of `r n_of_artefacts` final-stage spheroids and sub-spheroids from the site were included in the analysis.

@@

We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_of_artefacts`) and all other multifacial and polyhedron cores (n = ) in the assemblage of QSY. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core’s prototype, initial and final shapes, and cortex coverage here.

Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedron. The mean sphericity of spheroids (0.906) is notably higher than that of multifacial cores and polyhedron (0.861), with a Mann-Whitney U test yielding U = 37.0, p = 3.99 × 10⁻⁵. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.

@@

We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded H = 18.43, p = 3.59 × 10⁻⁴. The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.

The spherical harmonic energy (SHE), a novel metric introduced herein, quantifies the geometric approximation of core morphology to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (t = -11.012, df = 49, p = 7.443×10⁻¹⁵), validating its efficacy as a proxy of sphericity.

Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE=1.013 (SD ± 0.008), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE=1.032, SD ± 0.014). The Kruskal-Wallis test confirms robust inter-group divergence (H=16.99, p=3.76×10⁻⁵), proving the distinguishability of spheroids in QSY.

The pursuit of a standardized spherical form among spheroids is further supported by our findings. Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids. Eerkens (2000) suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts. Both sphericity and SHE values yielded low CVs (0.0175 and 0.0068, respectively), indicating that spheroids maintain a highly standardized rounded shape. In contrast, multifacial cores combined with polyhedron has a higher CV of 0.0447 and 0.0191 respectively.

@@

>>>>

```{r}
# 读取表格
library(tidyverse)
library(readxl)
library(here)
library(stringr)

file_path <- here("analysis/data/derived_data/core_typology_morphological_parameters.xlsx")
data_raw <- read_excel(file_path)

# 对所有字符型列进行清洗
data_clean <- data_raw |>
  mutate(across(
    where(is.character),
    ~ str_to_title(str_squish(trimws(.x)))
  ))

# 统计每类 Typology 的数量
typology_counts <- data_clean |> 
  count(Typology, sort = TRUE)

n_total_typology <- data_clean |> 
  filter(!is.na(Typology)) |> 
  nrow()

# 组合统计
group_counts <- data_clean |>
  filter(Typology %in% c("Polyhedron", "Multifacial", "Spheroid", "Subspheroid")) |>
  mutate(
    Group = case_when(
      Typology %in% c("Polyhedron", "Multifacial") ~ "Polyhedron + Multifacial",
      Typology %in% c("Spheroid", "Subspheroid") ~ "Spheroid + Subspheroid"
    )
  ) |>
  count(Group)

n_poly_multi <- group_counts$n[group_counts$Group == "Polyhedron + Multifacial"]
n_sub_spheroid <- group_counts$n[group_counts$Group == "Spheroid + Subspheroid"]
n_poly_multi

Based on typological study, we identify some faceted sub-spheroids/spheroids in QSY site. Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers. \[BM: feels like this belongs in results section\]

We applied quantitative methods to help confirm the degree of difference between the archaeological spheroids and a perfect sphere. This included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artefact. We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of sphericity and SHE.

@@

Where VVV is the volume and AAA is the surface area of the object. This dimensionless metric yields a maximum value of 1 for a perfect sphere. While values closer to 1 generally indicate a more spherical geometry, sphericity is calculated solely from volume and surface area. As a result, some irregular objects may exhibit high sphericity, while conversely, objects that are generally spherical in shape may deviate from ideal sphericity due to angular features or surface roughness. Thus, sphericity serves as a useful but limited proxy for assessing how close an object is to a perfect sphere. We used Python to derive surface area and volume measurements from 3D mesh models, and before that all models were decimated to a consistent face count and subjected to unit normalization to ensure comparability across artifacts.

@@

Spherical Harmonic (SPHARM) analysis is a method for describing 3D shapes using a series of mathematical functions defined on the surface of a sphere. It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies. Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$. The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes. Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively. Lower-order terms describe the general shape, while higher-order terms capture finer details. This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025]. Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.

We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020]. Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020]. The resulting coefficients were normalized for scale-invariant comparison across models.

@@

### Spherical Harmonic Energy

@@

To further investigate technological strategies, we incorporated orientation statistics as a complementary approach. This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts. We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps are finished in Geomagic Wrap, Rhino8 and R.

@@

Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies. We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes. This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution. A total of `r n_sub_spheroid` final-stage spheroids and sub-spheroids from the site were included in the analysis.

@@

We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the assemblage of QSY. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

```{r tbl-1, echo=FALSE}

knitr::kable(data_raw, caption = "Summary of core typology and morphological parameters.")

```{r}

# 统一 Typology 到两大组
mean_sphericity_tbl <- data_clean |>
  filter(Typology %in% c("Polyhedron", "Multifacial", "Spheroid", "Subspheroid")) |>
  mutate(
    Group = case_when(
      Typology %in% c("Polyhedron", "Multifacial") ~ "Polyhedron + Multifacial",
      Typology %in% c("Spheroid", "Subspheroid") ~ "Spheroid + Subspheroid"
    )
  )

# 计算每组的平均 Sphericity
mean_sphericity_summary <- mean_sphericity_tbl |>
  group_by(Group) |>
  summarise(mean_sphericity = mean(Sphericity, na.rm = TRUE)) |>
  ungroup()

# 提取每组的平均值，赋值为变量以便 qmd 引用
mean_sphericity_poly_multi <- mean_sphericity_summary$mean_sphericity[mean_sphericity_summary$Group == "Polyhedron + Multifacial"] |> round(3)
mean_sphericity_sub_spheroid <- mean_sphericity_summary$mean_sphericity[mean_sphericity_summary$Group == "Spheroid + Subspheroid"] |> round(3)

# Mann-Whitney U test

# 提取两组球度数据
group_poly_multi <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Polyhedron + Multifacial"]
group_sub_spheroid <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Spheroid + Subspheroid"]

mann_whitney_result_sphericity <- wilcox.test(group_poly_multi, group_sub_spheroid, exact = FALSE)

U_value <- mann_whitney_result_sphericity$statistic  # U 值
p_value <- mann_whitney_result_sphericity$p.value    # p 值

pretty_print_sci <- function(num) {
  sci_str <- sprintf("%.2e", num)
  parts <- strsplit(sci_str, "e")[[1]]
  base <- parts[1]
  exponent <- as.integer(parts[2])
  paste0(base, " \\times 10^{", exponent, "}")
}

p_value_Mann_sci <- pretty_print_sci(p_value)

# Kruskal–Wallis test
kruskal_result <- kruskal.test(Sphericity ~ Typology, data = data_clean)

H_value <- kruskal_result$statistic
df_kruskal <- kruskal_result$parameter
p_value_kruskal <- kruskal_result$p.value
p_value_kruskal_sci <- pretty_print_sci(p_value_kruskal)

# Pearson correlation test

cor_test_pearson <- cor.test(data_clean$Sphericity, data_clean$`SPHARM power`, method = "pearson")

# 提取结果
t_value <- cor_test_pearson$statistic
df_value <- cor_test_pearson$parameter
p_value <- cor_test_pearson$p.value
correlation <- cor_test_pearson$estimate

p_value_pearson_sci <- pretty_print_sci(p_value)

# Mean SHE

mean_spharm_power_summary <- mean_sphericity_tbl |>
  group_by(Group) |>
  summarise(
    mean_spharm_power = mean(`SPHARM power`, na.rm = TRUE),
    sd_spharm_power = sd(`SPHARM power`, na.rm = TRUE)
  ) |>
  ungroup()

mean_spharm_power_poly_multi <- mean_spharm_power_summary$mean_spharm_power[mean_spharm_power_summary$Group == "Polyhedron + Multifacial"] |> round(3)
sd_spharm_power_poly_multi <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Polyhedron + Multifacial"] |> round(3)

mean_spharm_power_sub_spheroid <- mean_spharm_power_summary$mean_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)
sd_spharm_power_sub_spheroid <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)

# Kruskal–Wallis test2
kruskal_result2 <- kruskal.test(`SPHARM power` ~ Typology, data = data_clean)

# 提取统计量和p值
H_value_kruskal2 <- kruskal_result2$statistic
df_kruskal2 <- kruskal_result2$parameter
p_value_kruskal2 <- kruskal_result2$p.value

p_value_kruskal2_sci <- pretty_print_sci(p_value_kruskal2)

# CV of sphericity and SHE

group_sub_spheroids <- mean_sphericity_tbl %>% filter(Group == "Spheroid + Subspheroid")

# 计算Sphericity的均值和标准差
mean_sph_sub_spheroids <- mean(group_sub_spheroids$Sphericity, na.rm = TRUE)
sd_sph_sub_spheroids <- sd(group_sub_spheroids$Sphericity, na.rm = TRUE)
cv_sph_sub_spheroids <- sd_sph_sub_spheroids / mean_sph_sub_spheroids

# 计算SHE的均值和标准差
mean_she_sub_spheroids <- mean(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
sd_she_sub_spheroids <- sd(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
cv_she_sub_spheroids <- sd_she_sub_spheroids / mean_she_sub_spheroids

group_poly_multi <- mean_sphericity_tbl %>% filter(Group == "Polyhedron + Multifacial")

# 计算Sphericity的均值和标准差
mean_sph_poly_multi <- mean(group_poly_multi$Sphericity, na.rm = TRUE)
sd_sph_poly_multi <- sd(group_poly_multi$Sphericity, na.rm = TRUE)
cv_sph_poly_multi <- sd_sph_poly_multi / mean_sph_poly_multi

# 计算SPHARM power的均值和标准差
mean_she_poly_multi <- mean(group_poly_multi$`SPHARM power`, na.rm = TRUE)
sd_she_poly_multi <- sd(group_poly_multi$`SPHARM power`, na.rm = TRUE)
cv_she_poly_multi <- sd_she_poly_multi / mean_she_poly_multi

cv_she_sub_spheroids
cv_she_poly_multi
```

Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra. The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.

@@

We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded $\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$. The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.

The spherical harmonic energy (SHE), a novel metric introduced herein, quantifies the geometric approximation of core morphology to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity.

Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`). The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids in QSY.

The pursuit of a standardized spherical form among spheroids is further supported by our findings. Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids. @Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts. Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that sub/spheroids maintain a highly standardized rounded shape. In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.

@@

## Efficiency of shaping spheroids
